---
title: "Lipidomics Analysis"
author: "Andrés Gordo Ortiz (PSancho Lab)"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    code_folding: hide
    code_download: false  
    df_print: paged
    theme: flatly
    highlight: tango
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
    number_sections: true
---

```{r logo, echo=FALSE}
htmltools::img(src = knitr::image_uri("logo.png"), 
               alt = 'logo', 
               style = 'position:absolute; top:0; right:0; width:150px; padding:10px;')

```

```{css, echo=FALSE}
/* Custom CSS to enhance styling */
body {
  font-family: 'Gill Sans', sans-serif;
}

h1 {
  color: #2c3e50;
  font-family: 'Ubuntu', sans-serif; /* Set header font to Ubuntu */
}

h2 {
  color: #3498db;
  font-family: 'Ubuntu', sans-serif; /* Set header font to Ubuntu */
}

h3 {
  color: #E888B4;
  font-family: 'Ubuntu', sans-serif; /* Set header font to Ubuntu */
}
p {
  text-align: justify;
}
```

```{r setup, include=FALSE}
library(ggplot2)
library(ggthemes)
library(plotly)
library(limma)
library(pheatmap)
library(tidyverse)
library(readxl)
library(lipidr)
library(DT)
library(limma)
library(cowplot)
library(svglite)
library(RColorBrewer) 
library(gplots)
library(heatmaply)
library(ggvenn)

source('ggplot_theme_Publication.R')
dir.create("plots", showWarnings = FALSE, recursive = TRUE)
dir.create("data", showWarnings = FALSE, recursive = TRUE)

knitr::opts_chunk$set(echo=TRUE, warning=FALSE, message=FALSE)
```

# Data Tidying
```{r tidydata}
data_253 <- read_excel("Copia de in vitro_mzml-out_5_7_1000-1.xlsx",sheet = "253")
#Delete every column but 1,2 and 22:30
data_253 <- data_253[,c(1,22:30)]
#Delete every row with NAs
data_253 <- data_253[complete.cases(data_253),]

#Repeat everything but with sheet="215"
data_215 <- read_excel("Copia de in vitro_mzml-out_5_7_1000-1.xlsx",sheet = "215")
data_215 <- data_215[,c(1,22:30)]
data_215 <- data_215[complete.cases(data_215),]

#Reorder data_253 to match data_215[[1]]
data_253 <- data_253[match(data_215[[1]],data_253[[1]]),]
data_253[,1]<-data_215[,1]

#Create a new merged dataframe from the two
data_253_215 <- merge(data_253,data_215,by="NAME",all=TRUE)


#Prepare sample names data_253_215 column names
sample_names<-names(data_253_215[,-1])

#Simplify data samples to Sx
names(data_253_215)<-c("NAME",paste0("S",1:18))

#Make Clinical (phenotypic excel)
data_clin_invitro<-data.frame(sample=names(data_253_215[,-1]), cell_line=as.factor(rep(c("line_253","line_215"),each=9)), 
                      treatment=rep(c("control", "eto", "mcm"), each=3, times=2),replica=as.factor(rep(c("rep_1","rep_2","rep_3"),times=6)),stringsAsFactors = FALSE)


#Eliminate parenthesis from data_253_215[,1]
data_matrix_invitro<-data_253_215
data_matrix_invitro[,1]<-gsub("\\(|\\)","",data_253_215[,1])
#Add a space between the last letter and first number of data_matrix[,1]
data_matrix_invitro[,1]<-gsub("(?<=[A-Za-z])(?=[0-9])"," ",data_matrix_invitro[,1],perl=TRUE)

#Delete all "-" in "-O" in data_matrix[,1]
data_matrix_invitro[,1]<-gsub("-O","O",data_matrix_invitro[,1])

#Change - symbol in some fatty acid names to make it standard
data_matrix_invitro[79,1]<-"DiHexCerd 18:1/14:13"
data_matrix_invitro[82,1]<-"DiHexCerd 18:1/16:15"
data_matrix_invitro[87,1]<-"DiHexCerd 18:1/22:21"
data_matrix_invitro[89,1]<-"DiHexCerd 18:1/24:23"
data_matrix_invitro[93,1]<-"DiHexCerd 20:1/12:11"


#Subset both cell lines
data_matrix_253<-data_matrix_invitro[, 1:10]
#Eliminate all rows with NAs in data_matrix_253
data_matrix_253<-data_matrix_253[complete.cases(data_matrix_253),]

data_matrix_215<-data_matrix_invitro[, c(1,11:19)]
#Eliminate all rows with NAs in data_matrix_253
data_matrix_215<-data_matrix_215[complete.cases(data_matrix_215),]

#Export data_clin and data_253_215 to .csv, without rownames
write.csv(data_clin_invitro,"data/data_clin_invitro.csv", row.names = FALSE)
write.csv(data_matrix_invitro,"data/data_matrix_invitro.csv", row.names = FALSE)


d_invitro <- as_lipidomics_experiment(read.csv("data_matrix_invitro.csv"))
d_invitro <- add_sample_annotation(d_invitro, "data_clin_invitro.csv")
```



# Quality Checking of Lipids

## Sample Quality
```{r qc_plot_samples}
#Quality check samples
qc_plot<-plot_samples(d_invitro, type = "boxplot", log = TRUE, color = "treatment")
qc_plot<-qc_plot +
  theme_Publication() +  # Change the theme to minimal
  labs(title = "QC Boxplot: Distribution of Log2-Area per sample") 
qc_plot
save_plot("plots/qc_plot_samples.png",qc_plot, base_width = 15, base_height = 7)

```

## Lipid Class Distribution

```{r qc_plot_lipid}
#Lipid class abundance
lipid_class_plot<-plot_lipidclass(d_invitro, "boxplot")

lipid_class_plot +
  theme_Publication() +
  labs(title = "Lipid Class Boxplot: Distribution of Log2-Area per Lipid subtype") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Angle the x-axis labels

save_plot("plots/qc_plot_lipids.png",lipid_class_plot +
  theme_Publication() +
  labs(title = "Lipid Class Boxplot: Distribution of Log2-Area per Lipid subtype") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)), base_width = 15, base_height = 7)

```

------------------------------------------------------------------------

# Principal Component Analysis

## All Samples (N253 & N215) {.tabset}

### By Cell Line
```{r pca_all_cellline}
exp_matrix<-assay(d_invitro)
exp_matrix_complete<-exp_matrix[complete.cases(exp_matrix),]
condition <- colData(d_invitro)$treatment
condition <- factor(condition)
sampleLabels<-rownames(colData(d_invitro))
cell_line<-colData(d_invitro)$cell_line

pca.res <- prcomp(t(exp_matrix_complete))
pc.var<-pca.res$sdev^2 # sdev^2 captures these eigenvalues from the PCA result
pc.per<-round(pc.var/sum(pc.var)*100, 1) 
pca.res.df <- pca.res$x[,1:4] %>% 
  as_tibble() %>%
  add_column(sample = sampleLabels,
             group = condition,
             cell_line= cell_line)

pca.plot_cell_line<- ggplot(pca.res.df) +
  aes(x=PC1, y=PC2, label=sampleLabels, color = cell_line) +
  geom_point(size=4) +
  stat_ellipse() +
  xlab(paste0("PC1 (",pc.per[1],"%",")")) + 
  ylab(paste0("PC2 (",pc.per[2],"%",")")) +
  labs(title="PCA plot coloured by Cell Line",
       caption=paste0("produced on ", Sys.time())) +
  coord_fixed() +
  theme_Publication()
pca.plot_cell_line
save_plot("plots/pca_all_samples_cell_line.png",pca.plot_cell_line, base_width = 15, base_height = 7)

```

### By Treatment
```{r pca_all_treatment}
pca.plot_treatment<- ggplot(pca.res.df) +
  aes(x=PC1, y=PC2, label=sampleLabels, color = condition) +
  geom_point(size=4) +
  stat_ellipse() +
  xlab(paste0("PC1 (",pc.per[1],"%",")")) + 
  ylab(paste0("PC2 (",pc.per[2],"%",")")) +
  labs(title="PCA plot coloured by Treatment",
       caption=paste0("produced on ", Sys.time())) +
  coord_fixed() +
  theme_Publication()
pca.plot_treatment
save_plot("plots/pca_all_samples_treatment.png",pca.plot_treatment, base_width = 15, base_height = 7)

```

### By Replicate
```{r pca_all_replicate}
replicate<-colData(d_invitro)$replica
pca.plot_replicate<- ggplot(pca.res.df) +
  aes(x=PC1, y=PC2, label=sampleLabels, color = replicate) +
  geom_point(size=4) +
  stat_ellipse() +
  xlab(paste0("PC1 (",pc.per[1],"%",")")) + 
  ylab(paste0("PC2 (",pc.per[2],"%",")")) +
  labs(title="PCA plot coloured by Condition",
       caption=paste0("produced on ", Sys.time())) +
  coord_fixed() +
  theme_Publication()
pca.plot_replicate
save_plot("plots/pca_all_samples_replicate.png",pca.plot_replicate, base_width = 15, base_height = 7)

```



### Small Multiples Plot
```{r smallmultiples_all}
pca.pivot <- pivot_longer(pca.res.df, 
                          cols = PC1:PC4, 
                          names_to = "PC", 
                          values_to = "loadings") 

small_multiples_plot <- ggplot(pca.pivot) +
  aes(x = sample, y = loadings, fill = sample) + 
  geom_bar(stat = "identity") +
  facet_wrap(~PC) +
  labs(title = "PCA 'small multiples' plot",
       caption = paste0("produced on ", Sys.time())) +
  theme_bw() +
  theme(legend.position = "none",
        axis.title.y = element_blank()) + 
  coord_flip()
small_multiples_plot
save_plot("plots/small_multiples_plot_all_samples.png",small_multiples_plot, base_width = 15, base_height = 7)
```

------------------------------------------------------------------------

## N253 Cell Line only {.tabset}

### By Treatment
```{r pca_N253_treatment}
exp_matrix<-assay(d_invitro)
exp_matrix_complete<-exp_matrix[complete.cases(exp_matrix),]
exp_matrix_complete_N253<-exp_matrix_complete[,1:9]
condition <- colData(d_invitro)$treatment[1:9]
condition <- factor(condition)
sampleLabels<-rownames(colData(d_invitro))[1:9]
cell_line<-colData(d_invitro)$cell_line[1:9]


pca.res <- prcomp(t(exp_matrix_complete_N253))
pc.var<-pca.res$sdev^2 # sdev^2 captures these eigenvalues from the PCA result
pc.per<-round(pc.var/sum(pc.var)*100, 1) 
pca.res.df <- pca.res$x[,1:4] %>% 
  as_tibble() %>%
  add_column(sample = sampleLabels,
             group = condition,
             cell_line= cell_line)

pca.plot_treatment<- ggplot(pca.res.df) +
  aes(x=PC1, y=PC2, label=sampleLabels, color = condition) +
  geom_point(size=4) +
  stat_ellipse() +
  xlab(paste0("PC1 (",pc.per[1],"%",")")) + 
  ylab(paste0("PC2 (",pc.per[2],"%",")")) +
  labs(title="PCA plot coloured by Condition in N253",
       caption=paste0("produced on ", Sys.time())) +
  coord_fixed() +
  theme_Publication()
pca.plot_treatment
save_plot("plots/pca_N253_treatment.png",pca.plot_treatment, base_width = 15, base_height = 7)
```

### Small Multiples Plot
```{r smallmultiples_N253}
pca.pivot <- pivot_longer(pca.res.df, 
                          cols = PC1:PC4, 
                          names_to = "PC", 
                          values_to = "loadings") 

small_multiples_plot <- ggplot(pca.pivot) +
  aes(x = sample, y = loadings, fill = sample) + 
  geom_bar(stat = "identity") +
  facet_wrap(~PC) +
  labs(title = "PCA 'small multiples' plot in N253",
       caption = paste0("produced on ", Sys.time())) +
  theme_bw() +
  theme(legend.position = "none",
        axis.title.y = element_blank()) + 
  coord_flip()
small_multiples_plot
save_plot("plots/small_multiples_plot_N253.png",small_multiples_plot, base_width = 15, base_height = 7)
```

## N215 Cell Line only {.tabset}

### By Treatment
```{r pca_N215_treatment}
exp_matrix<-assay(d_invitro)
exp_matrix_complete<-exp_matrix[complete.cases(exp_matrix),]
exp_matrix_complete_N215<-exp_matrix_complete[,10:18]
condition <- colData(d_invitro)$treatment[10:18]
condition <- factor(condition)
sampleLabels<-rownames(colData(d_invitro))[10:18]
cell_line<-colData(d_invitro)$cell_line[10:18]


pca.res <- prcomp(t(exp_matrix_complete_N215))
pc.var<-pca.res$sdev^2 # sdev^2 captures these eigenvalues from the PCA result
pc.per<-round(pc.var/sum(pc.var)*100, 1) 
pca.res.df <- pca.res$x[,1:4] %>% 
  as_tibble() %>%
  add_column(sample = sampleLabels,
             group = condition,
             cell_line= cell_line)

pca.plot_treatment<- ggplot(pca.res.df) +
  aes(x=PC1, y=PC2, label=sampleLabels, color = condition) +
  geom_point(size=4) +
  stat_ellipse() +
  xlab(paste0("PC1 (",pc.per[1],"%",")")) + 
  ylab(paste0("PC2 (",pc.per[2],"%",")")) +
  labs(title="PCA plot coloured by Condition in N215",
       caption=paste0("produced on ", Sys.time())) +
  coord_fixed() +
  theme_Publication()
pca.plot_treatment
save_plot("plots/pca_N215_treatment.png",pca.plot_treatment, base_width = 15, base_height = 7)
```

### Small Multiples Plot
```{r smallmultiples_N215}
pca.pivot <- pivot_longer(pca.res.df, 
                          cols = PC1:PC4, 
                          names_to = "PC", 
                          values_to = "loadings") 

small_multiples_plot <- ggplot(pca.pivot) +
  aes(x = sample, y = loadings, fill = sample) + 
  geom_bar(stat = "identity") +
  facet_wrap(~PC) +
  labs(title = "PCA 'small multiples' plot in N215",
       caption = paste0("produced on ", Sys.time())) +
  theme_bw() +
  theme(legend.position = "none",
        axis.title.y = element_blank()) + 
  coord_flip()
small_multiples_plot
save_plot("plots/small_multiples_plot_N215.png",small_multiples_plot, base_width = 15, base_height = 7)
```

------------------------------------------------------------------------

## Top Lipids by Loading {.tabset}

### Eto vs Control
```{r top loadings eto vs control lipidr}
mvaresults = mva(d_invitro[,data_clin_invitro$treatment %in% c("control", "eto")], method = "OPLS-DA", group_col = "treatment")
top_loadings_plots<-plot_mva_loadings(mvaresults, color_by="Class", top.n=15)
top_loadings_plots +
  theme_Publication() + 
  labs(title = "Loadings/Influence of each Lipid in OPLS-DA Analysis", subtitle = "Eto vs Control") +
  guides(alpha = "none")
save_plot("plots/top_loadings_eto_vs_control.png",top_loadings_plots +
  theme_Publication() + 
  labs(title = "Loadings/Influence of each Lipid in OPLS-DA Analysis", subtitle = "Eto vs Control") +
  guides(alpha = "none"), base_width = 10, base_height = 7)
top<-top_lipids(mvaresults, top.n=100)
top_molecules<-top$Molecule
write.csv(top,"data/top_loadings_opls_eto_vs_control.csv", row.names = FALSE)
```

```{r top lipids eto vs control}
datatable(top[, c(2,7,8)], 
          extensions = c('KeyTable', "FixedHeader"), 
          caption = "Top 100 Lipids by Loading Relevance Eto vs Control",
          options = list(keys = TRUE, searchHighlight = TRUE, pageLength = 10, lengthMenu = c("10", "25", "50", "100")))
```

### MCM vs Control
```{r top loadings mcm vs control lipidr}
mvaresults = mva(d_invitro[,data_clin_invitro$treatment %in% c("control", "mcm")], method = "OPLS-DA", group_col = "treatment")
top_loadings_plots<-plot_mva_loadings(mvaresults, color_by="Class", top.n=15)
top_loadings_plots +
  theme_Publication() + 
  labs(title = "Loadings/Influence of each Lipid in OPLS-DA Analysis", subtitle = "MCM vs Control") +
  guides(alpha = "none")
save_plot("plots/top_loadings_mcm_vs_control.png",top_loadings_plots +
  theme_Publication() + 
  labs(title = "Loadings/Influence of each Lipid in OPLS-DA Analysis", subtitle = "MCM vs Control") +
  guides(alpha = "none"), base_width = 10, base_height = 7)
top<-top_lipids(mvaresults, top.n=100)
top_molecules<-top$Molecule
write.csv(top,"data/top_loadings_opls_mcm_vs_control.csv", row.names = FALSE)
```

```{r top lipids mcm vs control}
datatable(top[, c(2,7,8)], 
          extensions = c('KeyTable', "FixedHeader"), 
          caption = "Top 100 Lipids by Loading Relevance MCM vs Control",
          options = list(keys = TRUE, searchHighlight = TRUE, pageLength = 10, lengthMenu = c("10", "25", "50", "100")))
```

------------------------------------------------------------------------

# Differential Abundance of Lipids

## N215

### MCM vs Control
```{r limma and Vplot mcm vs control N215}
condition <- colData(d_invitro)$treatment[10:18]
condition <- factor(condition)
sampleLabels<-rownames(colData(d_invitro))[10:18]
design <- model.matrix(~0 + condition)
colnames(design) <- levels(condition)

lipid_matrix_norm <- voom(exp_matrix_complete_N215, design, plot = FALSE)
fit <- lmFit(lipid_matrix_norm, design)
contrast.matrix_control_vs_mcm <- makeContrasts(treatment = mcm - control,
                                 levels=design)

fits <- contrasts.fit(fit, contrast.matrix_control_vs_mcm)
ebFit <- eBayes(fits)
myTopHits <- topTable(ebFit, adjust ="BH", coef=1, number=40000, sort.by="logFC")



myTopHits.df <- myTopHits %>%
  as_tibble(rownames = "Lipid_Name")


write.csv(myTopHits.df, "data/stats_mcm_vs_control_N215.csv", row.names = FALSE)



vplot <- ggplot(myTopHits.df) +
  aes(y=-log10(adj.P.Val), x=logFC, text = paste("Symbol:", Lipid_Name)) +
  geom_point(size=2) +
  geom_hline(yintercept = -log10(0.05), linetype="longdash", colour="grey", size=1) +
  geom_vline(xintercept = 1, linetype="longdash", colour="#BE684D", size=1) +
  geom_vline(xintercept = -1, linetype="longdash", colour="#2C467A", size=1) +
  #annotate("rect", xmin = 1, xmax = 12, ymin = -log10(0.01), ymax = 7.5, alpha=.2, fill="#BE684D") +
  #annotate("rect", xmin = -1, xmax = -12, ymin = -log10(0.01), ymax = 7.5, alpha=.2, fill="#2C467A") +
  labs(title="Volcano plot [MCM vs Control] N215",
       caption=paste0("produced on ", Sys.time())) +
  scale_x_continuous(n.breaks = 8)+
  theme_Publication()

save_plot("plots/volcano_mcm_vs_control_N215.svg",vplot)
save_plot("plots/volcano_mcm_vs_control_N215.png",vplot)

# Only in html ggplotly(vplot)
ggplotly(vplot)


```

#### Table of DE Lipids

To identify differentially expressed lipids, precision weights were first applied to each lipid based on its mean-variance relationship using [VOOM](https://genomebiology.biomedcentral.com/articles/10.1186/gb-2014-15-2-r29). Then, linear modeling and bayesian stats were employed via [Limma](https://academic.oup.com/nar/article/43/7/e47/2414268) to find lipids that were over- or under-represented by **1-logfold or more, with a false-discovery rate (FDR) of 0.05**. Multiple testing was assessed through the Benjamini-Hochberg method.

```{r DEL mcm vs control N215}
results <- decideTests(ebFit, method="global", adjust.method="BH", p.value=0.05, lfc=1)
colnames(lipid_matrix_norm$E) <- sampleLabels
diffLipids <- lipid_matrix_norm$E[results[,1] !=0,]
diffLipids.df <- as_tibble(diffLipids, rownames = "Lipid_Names")

resultsDEL<-myTopHits.df[myTopHits.df$Lipid_Name %in% diffLipids.df$Lipid_Names,]
write.csv(resultsDEL, "data/diffLipids_mcm_vs_control_N215.csv", row.names = FALSE)
res_N215_mcm<-nrow(resultsDEL)
DEL_N215_mcm<-resultsDEL$Lipid_Name
```

```{r table DEL mcm vs control N215}
datatable(resultsDEL, 
          extensions = c('KeyTable', "FixedHeader"),
          caption="Diferentially Expressed Lipids in MCM vs Control in N215",
          options = list(keys = TRUE, 
                         searchHighlight = TRUE, 
                         pageLength = 10, 
                         lengthMenu = c("10", "25", "50", "100")))

```

### Eto vs Control
```{r limma and Vplot etoc vs control N215}

contrast.matrix_control_vs_eto <- makeContrasts(treatment = eto - control,
                                 levels=design)

fits <- contrasts.fit(fit, contrast.matrix_control_vs_eto)
ebFit <- eBayes(fits)
myTopHits <- topTable(ebFit, adjust ="BH", coef=1, number=40000, sort.by="logFC")



myTopHits.df <- myTopHits %>%
  as_tibble(rownames = "Lipid_Name")


write.csv(myTopHits.df, "data/stats_eto_vs_control_N215.csv", row.names = FALSE)



vplot <- ggplot(myTopHits.df) +
  aes(y=-log10(adj.P.Val), x=logFC, text = paste("Symbol:", Lipid_Name)) +
  geom_point(size=2) +
  geom_hline(yintercept = -log10(0.05), linetype="longdash", colour="grey", size=1) +
  geom_vline(xintercept = 1, linetype="longdash", colour="#BE684D", size=1) +
  geom_vline(xintercept = -1, linetype="longdash", colour="#2C467A", size=1) +
  #annotate("rect", xmin = 1, xmax = 12, ymin = -log10(0.01), ymax = 7.5, alpha=.2, fill="#BE684D") +
  #annotate("rect", xmin = -1, xmax = -12, ymin = -log10(0.01), ymax = 7.5, alpha=.2, fill="#2C467A") +
  labs(title="Volcano plot [Eto vs Control] N215",
       caption=paste0("produced on ", Sys.time())) +
  scale_x_continuous(n.breaks = 8)+
  theme_Publication()

save_plot("plots/volcano_eto_vs_control_N215.svg",vplot)
save_plot("plots/volcano_eto_vs_control_N215.png",vplot)

# Only in html ggplotly(vplot)
ggplotly(vplot)


```

#### Table of DE Lipids

To identify differentially expressed lipids, precision weights were first applied to each lipid based on its mean-variance relationship using [VOOM](https://genomebiology.biomedcentral.com/articles/10.1186/gb-2014-15-2-r29). Then, linear modeling and bayesian stats were employed via [Limma](https://academic.oup.com/nar/article/43/7/e47/2414268) to find lipids that were over- or under-represented by **1-logfold or more, with a false-discovery rate (FDR) of 0.05**. Multiple testing was assessed through the *Benjamini-Hochberg* method.

```{r DEL eto vs control N215}
results <- decideTests(ebFit, method="global", adjust.method="BH", p.value=0.05, lfc=1)
colnames(lipid_matrix_norm$E) <- sampleLabels
diffLipids <- lipid_matrix_norm$E[results[,1] !=0,]
diffLipids.df <- as_tibble(diffLipids, rownames = "Lipid_Names")

resultsDEL<-myTopHits.df[myTopHits.df$Lipid_Name %in% diffLipids.df$Lipid_Names,]
write.csv(resultsDEL, "data/diffLipids_eto_vs_control_N215.csv", row.names = FALSE)
res_N215_eto<-nrow(resultsDEL)
DEL_N215_eto<-resultsDEL$Lipid_Name
```

```{r table DEL eto vs control N215}
datatable(resultsDEL, 
          extensions = c('KeyTable', "FixedHeader"),
          caption="Differentially Expressed Lipids in Eto vs Control N215",
          options = list(keys = TRUE, 
                         searchHighlight = TRUE, 
                         pageLength = 10, 
                         lengthMenu = c("10", "25", "50", "100")))

```


## N253

### MCM vs Control
```{r limma and Vplot mcm vs control N253}
condition <- colData(d_invitro)$treatment[1:9]
condition <- factor(condition)
sampleLabels<-rownames(colData(d_invitro))[1:9]
design <- model.matrix(~0 + condition)
colnames(design) <- levels(condition)

lipid_matrix_norm <- voom(exp_matrix_complete_N253, design, plot = FALSE)
fit <- lmFit(lipid_matrix_norm, design)
contrast.matrix_control_vs_mcm <- makeContrasts(treatment = mcm - control,
                                 levels=design)

fits <- contrasts.fit(fit, contrast.matrix_control_vs_mcm)
ebFit <- eBayes(fits)
myTopHits <- topTable(ebFit, adjust ="BH", coef=1, number=40000, sort.by="logFC")



myTopHits.df <- myTopHits %>%
  as_tibble(rownames = "Lipid_Name")


write.csv(myTopHits.df, "data/stats_mcm_vs_control_N253.csv", row.names = FALSE)



vplot <- ggplot(myTopHits.df) +
  aes(y=-log10(adj.P.Val), x=logFC, text = paste("Symbol:", Lipid_Name)) +
  geom_point(size=2) +
  geom_hline(yintercept = -log10(0.05), linetype="longdash", colour="grey", size=1) +
  geom_vline(xintercept = 1, linetype="longdash", colour="#BE684D", size=1) +
  geom_vline(xintercept = -1, linetype="longdash", colour="#2C467A", size=1) +
  #annotate("rect", xmin = 1, xmax = 12, ymin = -log10(0.01), ymax = 7.5, alpha=.2, fill="#BE684D") +
  #annotate("rect", xmin = -1, xmax = -12, ymin = -log10(0.01), ymax = 7.5, alpha=.2, fill="#2C467A") +
  labs(title="Volcano plot [MCM vs Control] N253",
       caption=paste0("produced on ", Sys.time())) +
  scale_x_continuous(n.breaks = 8)+
  theme_Publication()

save_plot("plots/volcano_mcm_vs_control_N253.svg",vplot)
save_plot("plots/volcano_mcm_vs_control_N253.png",vplot)

# Only in html ggplotly(vplot)
ggplotly(vplot)


```

#### Table of DE Lipids

To identify differentially expressed lipids, precision weights were first applied to each lipid based on its mean-variance relationship using [VOOM](https://genomebiology.biomedcentral.com/articles/10.1186/gb-2014-15-2-r29). Then, linear modeling and bayesian stats were employed via [Limma](https://academic.oup.com/nar/article/43/7/e47/2414268) to find lipids that were over- or under-represented by **1-logfold or more, with a false-discovery rate (FDR) of 0.05**. Multiple testing was assessed through the Benjamini-Hochberg method.

```{r DEL mcm vs control N253}
results <- decideTests(ebFit, method="global", adjust.method="BH", p.value=0.05, lfc=1)
colnames(lipid_matrix_norm$E) <- sampleLabels
diffLipids <- lipid_matrix_norm$E[results[,1] !=0,]
diffLipids.df <- as_tibble(diffLipids, rownames = "Lipid_Names")

resultsDEL<-myTopHits.df[myTopHits.df$Lipid_Name %in% diffLipids.df$Lipid_Names,]
write.csv(resultsDEL, "data/diffLipids_mcm_vs_control_N253.csv", row.names = FALSE)
res_N253_mcm<-nrow(resultsDEL)
DEL_N253_mcm<-resultsDEL$Lipid_Name

```

```{r table DEL mcm vs control N253}
datatable(resultsDEL, 
          extensions = c('KeyTable', "FixedHeader"),
          caption="Diferentially Expressed Lipids in MCM vs Control in N253",
          options = list(keys = TRUE, 
                         searchHighlight = TRUE, 
                         pageLength = 10, 
                         lengthMenu = c("10", "25", "50", "100")))

```

### Eto vs Control
```{r limma and Vplot etoc vs control N253}

contrast.matrix_control_vs_eto <- makeContrasts(treatment = eto - control,
                                 levels=design)

fits <- contrasts.fit(fit, contrast.matrix_control_vs_eto)
ebFit <- eBayes(fits)
myTopHits <- topTable(ebFit, adjust ="BH", coef=1, number=40000, sort.by="logFC")



myTopHits.df <- myTopHits %>%
  as_tibble(rownames = "Lipid_Name")


write.csv(myTopHits.df, "data/stats_eto_vs_control_N253.csv", row.names = FALSE)



vplot <- ggplot(myTopHits.df) +
  aes(y=-log10(adj.P.Val), x=logFC, text = paste("Symbol:", Lipid_Name)) +
  geom_point(size=2) +
  geom_hline(yintercept = -log10(0.05), linetype="longdash", colour="grey", size=1) +
  geom_vline(xintercept = 1, linetype="longdash", colour="#BE684D", size=1) +
  geom_vline(xintercept = -1, linetype="longdash", colour="#2C467A", size=1) +
  #annotate("rect", xmin = 1, xmax = 12, ymin = -log10(0.01), ymax = 7.5, alpha=.2, fill="#BE684D") +
  #annotate("rect", xmin = -1, xmax = -12, ymin = -log10(0.01), ymax = 7.5, alpha=.2, fill="#2C467A") +
  labs(title="Volcano plot [Eto vs Control] N253",
       caption=paste0("produced on ", Sys.time())) +
  scale_x_continuous(n.breaks = 8)+
  theme_Publication()

save_plot("plots/volcano_eto_vs_control_N253.svg",vplot)
save_plot("plots/volcano_eto_vs_control_N253.png",vplot)

# Only in html ggplotly(vplot)
ggplotly(vplot)


```

#### Table of DE Lipids

To identify differentially expressed lipids, precision weights were first applied to each lipid based on its mean-variance relationship using [VOOM](https://genomebiology.biomedcentral.com/articles/10.1186/gb-2014-15-2-r29). Then, linear modeling and bayesian stats were employed via [Limma](https://academic.oup.com/nar/article/43/7/e47/2414268) to find lipids that were over- or under-represented by **1-logfold or more, with a false-discovery rate (FDR) of 0.05**. Multiple testing was assessed through the *Benjamini-Hochberg* method.

```{r DEL eto vs control N253}
results <- decideTests(ebFit, method="global", adjust.method="BH", p.value=0.05, lfc=1)
colnames(lipid_matrix_norm$E) <- sampleLabels
diffLipids <- lipid_matrix_norm$E[results[,1] !=0,]
diffLipids.df <- as_tibble(diffLipids, rownames = "Lipid_Names")

resultsDEL<-myTopHits.df[myTopHits.df$Lipid_Name %in% diffLipids.df$Lipid_Names,]
write.csv(resultsDEL, "data/diffLipids_eto_vs_control_N253.csv", row.names = FALSE)
res_N253_eto<-nrow(resultsDEL)
DEL_N253_eto<-resultsDEL$Lipid_Name
```

```{r table DEL eto vs control N253}
datatable(resultsDEL, 
          extensions = c('KeyTable', "FixedHeader"),
          caption="Differentially Expressed Lipids in Eto vs Control N253",
          options = list(keys = TRUE, 
                         searchHighlight = TRUE, 
                         pageLength = 10, 
                         lengthMenu = c("10", "25", "50", "100")))

```

## All Cell lines

### MCM vs Control
```{r limma and Vplot mcm vs control all}
condition <- colData(d_invitro)$treatment
condition <- factor(condition)
sampleLabels<-rownames(colData(d_invitro))
design <- model.matrix(~0 + condition)
colnames(design) <- levels(condition)

lipid_matrix_norm <- voom(exp_matrix_complete, design, plot = FALSE)
fit <- lmFit(lipid_matrix_norm, design)
contrast.matrix_control_vs_mcm <- makeContrasts(treatment = mcm - control,
                                 levels=design)

fits <- contrasts.fit(fit, contrast.matrix_control_vs_mcm)
ebFit <- eBayes(fits)
myTopHits <- topTable(ebFit, adjust ="BH", coef=1, number=40000, sort.by="logFC")



myTopHits.df <- myTopHits %>%
  as_tibble(rownames = "Lipid_Name")


write.csv(myTopHits.df, "data/stats_mcm_vs_control_all_lines.csv", row.names = FALSE)



vplot <- ggplot(myTopHits.df) +
  aes(y=-log10(adj.P.Val), x=logFC, text = paste("Symbol:", Lipid_Name)) +
  geom_point(size=2) +
  geom_hline(yintercept = -log10(0.05), linetype="longdash", colour="grey", size=1) +
  geom_vline(xintercept = 1, linetype="longdash", colour="#BE684D", size=1) +
  geom_vline(xintercept = -1, linetype="longdash", colour="#2C467A", size=1) +
  #annotate("rect", xmin = 1, xmax = 12, ymin = -log10(0.01), ymax = 7.5, alpha=.2, fill="#BE684D") +
  #annotate("rect", xmin = -1, xmax = -12, ymin = -log10(0.01), ymax = 7.5, alpha=.2, fill="#2C467A") +
  labs(title="Volcano plot [MCM vs Control] All Lines",
       caption=paste0("produced on ", Sys.time())) +
  scale_x_continuous(n.breaks = 8)+
  theme_Publication()

save_plot("plots/volcano_mcm_vs_control_all.svg",vplot)
save_plot("plots/volcano_mcm_vs_control_all.png",vplot)

# Only in html ggplotly(vplot)
ggplotly(vplot)


```

#### Table of DE Lipids

To identify differentially expressed lipids, precision weights were first applied to each lipid based on its mean-variance relationship using [VOOM](https://genomebiology.biomedcentral.com/articles/10.1186/gb-2014-15-2-r29). Then, linear modeling and bayesian stats were employed via [Limma](https://academic.oup.com/nar/article/43/7/e47/2414268) to find lipids that were over- or under-represented by **1-logfold or more, with a false-discovery rate (FDR) of 0.05**. Multiple testing was assessed through the Benjamini-Hochberg method.

```{r DEL mcm vs control all}
results <- decideTests(ebFit, method="global", adjust.method="BH", p.value=0.05, lfc=1)
colnames(lipid_matrix_norm$E) <- sampleLabels
diffLipids <- lipid_matrix_norm$E[results[,1] !=0,]
diffLipids.df <- as_tibble(diffLipids, rownames = "Lipid_Names")

resultsDEL<-myTopHits.df[myTopHits.df$Lipid_Name %in% diffLipids.df$Lipid_Names,]
write.csv(resultsDEL, "data/diffLipids_mcm_vs_control_all.csv", row.names = FALSE)
res_all_mcm<-nrow(resultsDEL)
DEL_all_mcm<-resultsDEL$Lipid_Name
```

```{r table DEL mcm vs control all}
datatable(resultsDEL, 
          extensions = c('KeyTable', "FixedHeader"),
          caption="Diferentially Expressed Lipids in MCM vs Control in All Lines",
          options = list(keys = TRUE, 
                         searchHighlight = TRUE, 
                         pageLength = 10, 
                         lengthMenu = c("10", "25", "50", "100")))

```

### Eto vs Control
```{r limma and Vplot etoc vs control all}

contrast.matrix_control_vs_eto <- makeContrasts(treatment = eto - control,
                                 levels=design)

fits <- contrasts.fit(fit, contrast.matrix_control_vs_eto)
ebFit <- eBayes(fits)
myTopHits <- topTable(ebFit, adjust ="BH", coef=1, number=40000, sort.by="logFC")



myTopHits.df <- myTopHits %>%
  as_tibble(rownames = "Lipid_Name")


write.csv(myTopHits.df, "data/stats_eto_vs_control_all.csv", row.names = FALSE)



vplot <- ggplot(myTopHits.df) +
  aes(y=-log10(adj.P.Val), x=logFC, text = paste("Symbol:", Lipid_Name)) +
  geom_point(size=2) +
  geom_hline(yintercept = -log10(0.05), linetype="longdash", colour="grey", size=1) +
  geom_vline(xintercept = 1, linetype="longdash", colour="#BE684D", size=1) +
  geom_vline(xintercept = -1, linetype="longdash", colour="#2C467A", size=1) +
  #annotate("rect", xmin = 1, xmax = 12, ymin = -log10(0.01), ymax = 7.5, alpha=.2, fill="#BE684D") +
  #annotate("rect", xmin = -1, xmax = -12, ymin = -log10(0.01), ymax = 7.5, alpha=.2, fill="#2C467A") +
  labs(title="Volcano plot [Eto vs Control] All Lines",
       caption=paste0("produced on ", Sys.time())) +
  scale_x_continuous(n.breaks = 8)+
  theme_Publication()

save_plot("plots/volcano_eto_vs_control_all.svg",vplot)
save_plot("plots/volcano_eto_vs_control_all.png",vplot)

# Only in html ggplotly(vplot)
ggplotly(vplot)


```

#### Table of DE Lipids

To identify differentially expressed lipids, precision weights were first applied to each lipid based on its mean-variance relationship using [VOOM](https://genomebiology.biomedcentral.com/articles/10.1186/gb-2014-15-2-r29). Then, linear modeling and bayesian stats were employed via [Limma](https://academic.oup.com/nar/article/43/7/e47/2414268) to find lipids that were over- or under-represented by **1-logfold or more, with a false-discovery rate (FDR) of 0.05**. Multiple testing was assessed through the *Benjamini-Hochberg* method.

```{r DEL eto vs control all}
results <- decideTests(ebFit, method="global", adjust.method="BH", p.value=0.05, lfc=1)
colnames(lipid_matrix_norm$E) <- sampleLabels
diffLipids <- lipid_matrix_norm$E[results[,1] !=0,]
diffLipids.df <- as_tibble(diffLipids, rownames = "Lipid_Names")

resultsDEL<-myTopHits.df[myTopHits.df$Lipid_Name %in% diffLipids.df$Lipid_Names,]
write.csv(resultsDEL, "data/diffLipids_eto_vs_control_all.csv", row.names = FALSE)
res_all_eto<-nrow(resultsDEL)
DEL_all_eto<-resultsDEL$Lipid_Name
```

```{r table DEL eto vs control all}
datatable(resultsDEL, 
          extensions = c('KeyTable', "FixedHeader"),
          caption="Differentially Expressed Lipids in Eto vs Control All Lines",
          options = list(keys = TRUE, 
                         searchHighlight = TRUE, 
                         pageLength = 10, 
                         lengthMenu = c("10", "25", "50", "100")))

```

## Statistical Power Effect

```{r stats plot }
stats_compare<-data.frame(del=c(res_all_eto, res_all_mcm, res_N253_eto, res_N253_mcm, res_N215_eto, res_N215_mcm), cell_line=rep(c("Both","N253", "N215"), each=2), comparison=rep(c("eto","mcm"),times=3))
# Plot a ggplot bar plot
ggplot(stats_compare, aes(x=cell_line, y=del, fill=comparison)) +
  geom_bar(stat="identity", position=position_dodge()) +
  labs(title="Number of DE Lipids in Each Comparison",
       y="Number of DE Lipids",
       x="Subset Used",
       caption=paste0("produced on ", Sys.time())) +
  theme_Publication() +
  scale_fill_manual(values=c("#2C467A", "#BE684D", "#F0E442")) +
  theme(legend.position = "top")

```

## Venn Diagram {.tabset}

### Eto vs Control
```{r venn eto}
venn_eto <- ggvenn(data = list(N215 = DEL_N215_eto, N253 = DEL_N253_eto, Both=DEL_all_eto))
venn_eto
```

### MCM vs Control
```{r venn mcm}
venn_mcm <- ggvenn(data = list(N215 = DEL_N215_mcm, N253 = DEL_N253_mcm, Both=DEL_all_mcm))
venn_mcm
```

------------------------------------------------------------------------

# Heatmaps {.tabset}

Pearson correlation was used to cluster all differentially expressed lipids, which were then represented as heatmap with the data scaled by *Zscore* for each row. On the other hand, Spearman correlation was used to cluster all samples. Two *modules* can be seen, being either over- or under-represented depending on the condition of the samples.

## Only DE Lipids Eto vs Control

```{r heatmap DEL, results='hide'}
myheatcolors <- rev(brewer.pal(name="RdBu", n=11))
clustRows <- hclust(as.dist(1-cor(t(diffLipids), method="pearson")), method="complete") 

clustColumns <- hclust(as.dist(1-cor(diffLipids, method="spearman")), method="complete")
module.assign <- cutree(clustRows, k=2) #Important parametre
module.color <- rainbow(length(unique(module.assign)), start=0.1, end=0.9) 
module.color <- module.color[as.vector(module.assign)] 

module.assign_col<-cutree(clustColumns, k=2) #Important parametre
module.color_col <- rainbow(length(unique(module.assign_col)), start=0.1, end=0.9) 
module.color_col <- module.color_col[as.vector(module.assign_col)]

svg(file = "plots/heatmap_DEL_eto_vs_control.svg")
heatmap.2(diffLipids, 
          Rowv = as.dendrogram(clustRows), 
          Colv = as.dendrogram(clustColumns),
          RowSideColors = module.color,
          ColSideColors = module.color_col,
          col = myheatcolors, 
          scale = "row", 
          labRow = NA,
          density.info = "none", 
          trace = "none",  
          cexRow = 0.8,  # Adjust row label size
          cexCol = 0.8,  # Adjust column label size
          margins = c(10, 10),  # Increase margin size
          keysize = 1.2,  # Adjust size of the color key
          key.title = NA,  # Remove color key title
          key.ylab = NULL,  # Remove color key y-axis label
          key.xlab = "Z-Score",  # Add x-axis label to the color key
          main = "Heatmap of Lipid Differences",  # Add a main title
          xlab = "Samples",  # Label for x-axis
          ylab = "Lipids",  # Label for y-axis
          mar = c(5, 4),  # Adjust plot margins
          srtCol = 45,  # Rotate column labels for better readability
          dendrogram = "both"  # Show both row and column dendrograms
)
dev.off()

png(file = "plots/heatmap_DEL_eto_vs_control.png")
heatmap.2(diffLipids, 
          Rowv = as.dendrogram(clustRows), 
          Colv = as.dendrogram(clustColumns),
          RowSideColors = module.color,
          ColSideColors = module.color_col,
          col = myheatcolors, 
          scale = "row", 
          labRow = NA,
          density.info = "none", 
          trace = "none",  
          cexRow = 0.8,  # Adjust row label size
          cexCol = 0.8,  # Adjust column label size
          margins = c(10, 10),  # Increase margin size
          keysize = 1.2,  # Adjust size of the color key
          key.title = NA,  # Remove color key title
          key.ylab = NULL,  # Remove color key y-axis label
          key.xlab = "Z-Score",  # Add x-axis label to the color key
          main = "Heatmap of Lipid Differences",  # Add a main title
          xlab = "Samples",  # Label for x-axis
          ylab = "Lipids",  # Label for y-axis
          mar = c(5, 4),  # Adjust plot margins
          srtCol = 45,  # Rotate column labels for better readability
          dendrogram = "both"  # Show both row and column dendrograms
)
dev.off()
```

```{r heatmaply DELs}

heatmaply(diffLipids, 
          colors = myheatcolors,
          Rowv=as.dendrogram(clustRows),
          RowSideColors=module.color,
          ColSideColors = module.color_col,
          showticklabels=c(TRUE,FALSE),
          scale='row',
          main ="Heatmap across all Lipids")

```

## All Lipids

```{r heatmap ALL, results='hide'}
myheatcolors <- rev(brewer.pal(name="RdBu", n=11))
clustRows <- hclust(as.dist(1-cor(t(lipid_matrix_norm$E), method="pearson")), method="complete") 

clustColumns <- hclust(as.dist(1-cor(lipid_matrix_norm$E, method="spearman")), method="complete")
module.assign <- cutree(clustRows, k=2) #Important parametre
module.color <- rainbow(length(unique(module.assign)), start=0.1, end=0.9) 
module.color <- module.color[as.vector(module.assign)] 

module.assign_col<-cutree(clustColumns, k=2) #Important parametre
module.color_col <- rainbow(length(unique(module.assign_col)), start=0.1, end=0.9) 
module.color_col <- module.color_col[as.vector(module.assign_col)]

svg(file = "plots/heatmap_all.svg")
heatmap.2(lipid_matrix_norm$E, 
          Rowv = as.dendrogram(clustRows), 
          Colv = as.dendrogram(clustColumns),
          RowSideColors = module.color,
          ColSideColors = module.color_col,
          col = myheatcolors, 
          scale = "row", 
          labRow = NA,
          density.info = "none", 
          trace = "none",  
          cexRow = 0.8,  # Adjust row label size
          cexCol = 0.8,  # Adjust column label size
          margins = c(10, 10),  # Increase margin size
          keysize = 1.2,  # Adjust size of the color key
          key.title = NA,  # Remove color key title
          key.ylab = NULL,  # Remove color key y-axis label
          key.xlab = "Z-Score",  # Add x-axis label to the color key
          main = "Heatmap of Lipid Differences",  # Add a main title
          xlab = "Samples",  # Label for x-axis
          ylab = "Lipids",  # Label for y-axis
          mar = c(5, 4),  # Adjust plot margins
          srtCol = 45,  # Rotate column labels for better readability
          dendrogram = "both"  # Show both row and column dendrograms
)
dev.off()

png(file = "plots/heatmap_all.png")
heatmap.2(lipid_matrix_norm$E, 
          Rowv = as.dendrogram(clustRows), 
          Colv = as.dendrogram(clustColumns),
          RowSideColors = module.color,
          ColSideColors = module.color_col,
          col = myheatcolors, 
          scale = "row", 
          labRow = NA,
          density.info = "none", 
          trace = "none",  
          cexRow = 0.8,  # Adjust row label size
          cexCol = 0.8,  # Adjust column label size
          margins = c(10, 10),  # Increase margin size
          keysize = 1.2,  # Adjust size of the color key
          key.title = NA,  # Remove color key title
          key.ylab = NULL,  # Remove color key y-axis label
          key.xlab = "Z-Score",  # Add x-axis label to the color key
          main = "Heatmap of Lipid Differences",  # Add a main title
          xlab = "Samples",  # Label for x-axis
          ylab = "Lipids",  # Label for y-axis
          mar = c(5, 4),  # Adjust plot margins
          srtCol = 45,  # Rotate column labels for better readability
          dendrogram = "both"  # Show both row and column dendrograms
)
dev.off()
```

```{r heatmaply all}
heatmaply(lipid_matrix_norm$E, 
          colors = myheatcolors,
          Rowv=as.dendrogram(clustRows),
          RowSideColors=module.color,
          ColSideColors = module.color_col,
          showticklabels=c(TRUE,FALSE),
          scale='row',
          main ="Heatmap across all Lipids")
```


------------------------------------------------------------------------



# Session info

The output from running 'sessionInfo' is shown below and details all packages and version necessary to reproduce the results in this report.

```{r session info}
sessionInfo()
```